Introduction to Markov state modeling with the PyEMMA software --- v\versionnumber

### 2018-09-20 manuscript/manuscript.tex
addressing @psolsson's comments [ci skip]

The estimation of an MSM requires the dynamics between microstates to be Markovian.
However, in case of a poor discretization or short trajectories,
it might be hard or even impossible to meet this requirement (as we will see in Notebook~07).

An alternative, which is much less sensitive to poor discretization,
is to estimate a hidden Markov model (HMM)~\cite{hmm-baum-welch-alg,hmm-tutorial,jhp-spectral-rate-theory,bhmm-preprint}.
HMMs require the dynamics in some hidden (i.e., unobserved) state space to be Markovian.

To estimate an HMM, we need a spectral gap after the $m^\textrm{th}$ eigenvalue;
in practice, a timescale separation of $t_m \geq 2t_{m+1}$ is sufficient~\cite{pyemma}.
Then, we can approximate the dynamics in the observed microstates ($\mathbf{P}$) at any lag time $k\tau$ via
\begin{equation}
\mathbf{P}(k\tau) \approx \bm{\Pi}^{-1} \bm{\chi}^\top \tilde{\bm{\Pi}} \tilde{\mathbf{P}}^k(\tau) \, \bm{\chi}.
\end{equation}
Here, the $\bm{\Pi}=\left[ \pi_1,\dots,\pi_n \right]$ is a diagonal matrix of the $n$ microstates' stationary probabilities,
$\tilde{\bm{\Pi}}=\left[ \tilde{\pi}_1,\dots,\tilde{\pi}_m \right]$ is a diagonal matrix of the $m$ hidden states' stationary probabilities,
$\tilde{\mathbf{P}}(\tau)$ is a (small) transition matrix between the $m$ hidden states at lag time $\tau$,
and $\bm{\chi}$ is an $m\times n$-dimensional row-stochastic matrix
where each row encodes the probabilities of the $n$ microstates to appear in the corresponding hidden state~\cite{noe-proj-hid-msm}.

An HMM estimation usually yields a model with a small number of (hidden) states
where each state is considered to be metastable and,
thus, the number of hidden states is a new hyper-parameter which needs to be chosen carefully (see Notebook~07).
Given that an HMM is a generalization of an MSM,
we can compute all physico-chemically relevant quantities available for MSMs, too.

>>>>

The estimation of an MSM requires the dynamics between microstates to be Markovian.
However, in case of a poor dimension reduction and/or discretization or short trajectories,
we cannot anticipate this to be the case.
We illustrate this point in Notebook~07.

An alternative, which is much less sensitive to poor discretization,
is to estimate a hidden Markov model (HMM)~\cite{hmm-baum-welch-alg,hmm-tutorial,jhp-spectral-rate-theory,bhmm-preprint}.
HMMs are less sensitive to the discretization error as they sidestep the assumption of Markovian dynamics in the discretized space.
Instead, HMMs assume that there is an underlying (hidden) dynamic process which is Markovian
and gives rise to our observed data, e.g., the discretized trajectories.
This is a powerful principle as we know that there is indeed an underlying process which is Markovian:
our molecular dynamics trajectories.

To estimate an HMM, we need a spectral gap after the $m^\textrm{th}$ eigenvalue;
in practice, a timescale separation of $t_m \geq 2t_{m+1}$ is sufficient~\cite{pyemma}.
Then, we can approximate the dynamics in the observed microstates ($\mathbf{P}$) at any lag time $k\tau$ via
\begin{equation}
\mathbf{P}(k\tau) \approx \bm{\Pi}^{-1} \bm{\chi}^\top \tilde{\bm{\Pi}} \tilde{\mathbf{P}}^k(\tau) \, \bm{\chi}.
\end{equation}
Here, the $\bm{\Pi}=\left[ \pi_1,\dots,\pi_n \right]$ is a diagonal matrix of the $n$ microstates' stationary probabilities,
$\tilde{\bm{\Pi}}=\left[ \tilde{\pi}_1,\dots,\tilde{\pi}_m \right]$ is a diagonal matrix of the $m<n$ hidden states' stationary probabilities,
$\tilde{\mathbf{P}}(\tau)$ is a transition matrix between the $m<n$ hidden states at lag time $\tau$,
and $\bm{\chi}$ is an $m\times n$-dimensional row-stochastic matrix
where each row encodes the probabilities of the $n$ microstates to appear in the corresponding hidden state~\cite{noe-proj-hid-msm}.

An HMM estimation always yields a model with a small number of (hidden) states
where each state is considered to be metastable and,
thus, the number of hidden states is a new hyper-parameter which needs to be chosen carefully (see Notebook~07).
As the HMMs---like MSMs---approximate the full phase-space dynamics,
we can similarly compute physical observables from directly.
